\textsf{Are frontier models superhuman chemists?}

### 2024-03-12 src/tex/ms.tex
feat: update manuscript, fix analysis bugs

\section{Introduction}
\Glspl{llm} are frontier \gls{ml} models trained on massive amounts of text to complete sentences.
While some see in them \enquote{sparks of \gls{agi}},\cite{bubeck2023sparks} others consider them as \enquote{stochasitc parrots}---i.e., systems that only regurgitate what they have been trained on.\cite{bender2021dangers}

Chemists and materials scientists have quickly caught on the mounting attention given to \glspl{llm}, some even suggesting that \enquote{the future of chemistry is language}.\cite{White_2023}
This statement is motivated by a growing number of reports that use \glspl{llm} to properties of molecules or materials,\cite{jablonka202314, jablonka2024leveraging, xie2024fine} to optimize reactions\cite{ramos2023bayesian, kristiadi2024sober} or generate materials,\cite{rubungo2023llm, flam2023language, gruver2024fine} extract information,\cite{Patiny_2023, Dagdelen_2024, Zheng_2024, lála2023paperqa, caufield2023structured} or to even build \enquote{autonomous} systems that can physically perform reactions provided a command in natural language.\cite{bran2023chemcrow, Boiko_2023, darvish2024organa}
The rapid increase in capabilities led to concerns about the potential for dual use of these technologies, e.g., for the design of chemical weapons.\cite{gopal2023releasing, ganguli2022red, Urbina_2022}
Moreover, the use of these models is now also widespread among students\cite{Intelligent.com_2023} as well as research groups. For some users, misleading information---especially about safety-related aspects---might lead to harmful outcomes. Unfortunately, apart from anecdotal reports there is little evidence on how \glspl{llm} perform compared to experts.

Thus, to better understand what \glspl{llm} can do for chemistry and materials science, and where they might be improved with further developments, a comprehensive analysis is needed.
For the development of \glspl{llm}, such evaluation is currently mostly performed via standardized benchmarks such as BigBench\cite{srivastava2022beyond} or the LM Eval Harness.\cite{eval-harness}
The former contains, among 204 tasks, only two tasks classified as \enquote{chemistry related} whereas the latter contains no specific chemistry tasks.
Due to the lack of widely excepted standard benchmarks, the developers of chemical language models\cite{jablonka2024leveraging, guo2023large, ahmad2022chemberta2, Cai_2024} frequently utilize language-interfaced\cite{dinh2022lift} tabular datasets such as the ones reported in MoleculeNet,\cite{wu2018moleculenet} Therapeutic Data Commons\cite{huang2021therapeutics} or MatBench.\cite{dunn2020benchmarking}
While those evaluations can measure how well models can make predictions for very specific tasks, they only give a poor measure of how useful those models might be as a chemical assistant.

While some benchmark based on university entrance exams\cite{Zaki_2024, arora2023llms} or automatic text mining\cite{song2023honeybee, wei2021chemistryqa} have been proposed, also those do not satisfy the following basic criteria chemistry benchmarks should satisfy:
\begin{itemize}
    \item \emph{End-to-end automation}. For model development, the evaluations will need to be run many times (e.g., on regular intervals of a training run).
    Approaches that rely on humans scoring the answers of a system\cite{Schulze_Balhorn_2024, ai4science2023impact} can thus not be used.
    \item \emph{Careful validation by experts}. Manual curation is needed to minimize number of incorrect or unanswerable questions.\cite{northcutt2021pervasive}
    This is motivated by the observation that many widely used benchmarks are plagued by noisiness.\cite{Frye_2023, Awg}
    \item \emph{Usable with models that support special treatment of molecules}. Some models such as Galactica\cite{taylor2022galactica} use special tokenization or encoding procedures for molecules or equations.
    To support this, the benchmark system must encode the semantic meaning of various parts of the question or answer.
    \item \emph{Usable with black box systems}. Many relevant systems do not provide access to model weights or even just the raw logits.
    This might be the case because the systems are proprietary or because they involve not only \glspl{llm} but also external tools such as search \glspl{api} or code executors.\cite{schick2024toolformer, karpas2022mrkl, yao2022react}
    Thus, a benchmark should not assume access to the raw model outputs but be able to operate on text completions.
    \item \emph{Probing capabilities beyond answering of \glspl{mcq}}. In real world chemistry as well as higher-level university education multiple choice question are seldom utilized.
    Yet, most benchmarking frameworks focus on the \gls{mcq} setting because of the ease of evaluation. Realistic evaluations must measure capabilities beyond the answering of \gls{mcq}.
    \item \emph{Cover a diverse set of topics}. Chemistry, as the \enquote{central science}, bridges multiple disciplines.\cite{Aspuru_Guzik_2018} To even just approximate \enquote{chemistry capabilities} the topics covered by a chemistry benchmark must be very diverse.
\end{itemize}

In this work, we report a novel benchmarking framework, chembench, and use it to reveal limitations of current frontier models for the use in the chemical sciences.
Our benchmark consists of \variable{output/total_number_of_questions.txt}\unskip question answer pairs manually (\variable{output/manually_generated.txt}\unskip) or semi-automatically (\variable{output/automatically_generated.txt}\unskip) compiled from diverse sources.
It covers a large fraction of the topics taught in undergraduate chemistry curricula at various skill levels and can be used with any system that can return text (i.e., also tool-augmented systems).

To contextualize the scores, we also surveyed more than \variable{output/number_experts.txt} experts in chemistry (for a total of number of more than \variable{output/total_hours.txt}\unskip) on a subset of the benchmark corpus to be able to compare the performance of current frontier models with the one of humans.
Our results indicate that current frontier models perform \enquote{superhuman} on some aspects of chemistry but in many cases, included safety-related ones, might be very misleading.

@@

\subsection{Benchmark dataset}

To compile our benchmark corpus we utilized a broad list of sources (\Cref{sec:curation}), ranging from university exams to semi-automatically generated questions based on curated subsets of data in chemical databases.
To ensure maximal interoperability, we curated the data in an extended form of the BigBench format.
This also implies that future baselines can be built on top of our infrastructure as long as they are saved in the same format.
For quality assurance, all questions have been reviewed by at least one scientist in addition to the original curator.

Importantly, our large pool of questions encompassed a wide range of topics.
This can be seen, for example, in \Cref{fig:topic_barplot} in which we compare the number of questions in different categories (see \Cref{sec:meth-topic} for details on how we assigned topics).
By design, a focus of our corpus is on safety-related aspects with a (currently) limited sampling of questions from physical or theoretical chemistry, which might be extended in future work.

@@

Importantly, the corpus samples both \gls{mcq} and open-ended questions in a balanced way (XX \gls{mcq} questions vs. XX open-ended questions).
As one might expect, most questions are difficult to read according to the Flesch–Kincaid readability test (\variable{output/flesch_kincaid_reading_ease.txt}).~\cite{kincaid1975derivation}

\begin{figure}
    \centering
    \includegraphics{figures/wordcloud.pdf}
    \caption{\textbf{Wordcloud.}}
    \label{fig:wordcloud}
    \script{wordcloud.py}
\end{figure}

\subsection{\enquote{Tiny} subset}
It is important to note that for routine evaluations, a smaller subset of the corpus might be more practical.\cite{polo2024tinybenchmarks}
For instance,~\citet{liang2023holistic} report costs of more than 10,000 USD for \gls{api} calls for a single evaluation on the \gls{helm} benchmark.

@@

\subsection{Future work}
- Advanced prompting techniques such as CoT and test time strategies

- More advanced systems with chemnistry specific tools

- More systematic human baseline: Ask people to come in, do not allow to skip questions

@@

\subparagraph{Analytical chemistry}
The questions are sourced from master student examination papers at the University of Jena, Germany. These questions focus on the principles and techniques employed in analytical chemistry.

\subparagraph{Combustion engineering}
These questions are derived from master student examination papers at the University of Magdeburg, Germany. They explore topics related to combustion processes and engineering principles.

\subparagraph{Functional materials and nano-materials}
The questions are gathered from a seminar course conducted at the University of Jena, delving into topics concerning functional materials and nano-materials, exploring their properties and applications.

\subparagraph{General chemistry}
The questions originate from 10th grade examinations at Marist Comprehensive Academy Uturu, Nigeria. They cover fundamental concepts in chemistry suitable for students at the secondary school level.

\subparagraph{International Chemical Olympiad (\gls{ICHO})}
The questions are sourced from various countries' Chemical Olympiads, including those held in the USA, UK, and Moldova.

\subparagraph{Material synthesis}
These questions are derived from seminars on material synthesis at the University of Jena, focusing on the methods and techniques employed in the synthesis of various materials.

\subparagraph{Organic reactivity}
The questions are primarily sourced from tutorial papers from the Technical University of Munich, Germany, exploring reactions and mechanisms in organic chemistry.

\subparagraph{Periodic table}
The questions are manually created based on the reference periodic table from the Royal Society of Chemistry (\url{https://www.rsc.org/periodic-table/}), covering various aspects and trends of the periodic table.

\subparagraph{Polymer chemistry}
Questions are sourced from examinations held at the University of Hamburg, covering topics related to polymer chemistry and its applications.

\subparagraph{Chemical reactivity}
These questions are framed based on information from the Cameo Chemicals website (\url{https://cameochemicals.noaa.gov/reactivity}), exploring the reactivity of various chemicals and compounds.

\subparagraph{Chemical safety}
The questions encompass various aspects of safety, including lab safety, toxicology, material safety data sheets, carcinogens, sourced from reputable safety resources.

\subparagraph{Shape of molecules}
The questions are sourced from examinations held at the Technical University of Munich, Germany, focusing on determining the shape and geometry of molecules.

\subparagraph{\gls{NMR} spectroscopy}
Questions are sourced from the twitter page called (\url{https://twitter.com/NMRspectroscopy}), focusing on concepts and applications of nuclear magnetic resonance spectroscopy.

@@

\subparagraph{Oxidation states}
The questions regarding oxidation states are extracted from the website \url{https://www.cheminfo.org/}. These exercises focus on determining the number of isomers for specific molecules. To perform these calculations, we utilize the RDKit package \cite{bento2020open}.
\subparagraph{Electron count of atoms}
Questions pertaining to the electron count of atoms are gathered from the website \url{https://www.cheminfo.org/}. These inquiries involve determining the electron count of atoms in various chemical contexts. The calculations for these tasks are facilitated using the RDKit package \cite{bento2020open}.
\subparagraph{Point group for molecules}
Questions involve predicting the point group for molecules. The ChemCaption tool (\url{https://github.com/lamalab-org/chem-caption/tree/main}) is utilized for data generation. Given a molecule, predict its point group.
\subparagraph{SMILES-IUPAC}
Web scrapped SMILES IUPAC pairs from the PubChem database \cite{pubchem} and then framed questions involve converting SMILES notation to IUPAC nomenclature, utilizing information from the PubChem database.

@@

To maintain the consistency with the training, we employ distinct prompt templates in general tailored for completion models and instruction-tuned models. 
We pose instruction or impose constraints on the models within these templates to receive response in a specific format so that a robust, fair and consistent parsing can be performed as explained in \Graf{sec:parsing}.
Certain models are trained with special annotations and LaTeX syntax for scientific notations, chemical reactions, or symbols embedded within the text. 
For example, all the SMILES representations are encapsulated within \texttt{[START_SMILES][\text backslash END_SMILES]} in Galactica\cite{taylor2022galactica}.
Our prompting strategy consistently adheres to  these details in a model-specific manner by post-processing \latex syntax, chemical symbols, chemical equations and physical units (by either adding or removing wrappers).

\subparagraph{Completion models}
\texttt{Galactica-120b}\cite{taylor2022galactica}

@@

\paragraph{Models}

@@

\paragraph{Question selection}
Since we anticipated that we will not be able to collect enough responses for every question to allow for a meaningful statistical analysis, we decided on showing a relevant subset of all questions to the human scorers.
For selecting the subset, we decided on addressing two questions:
\begin{itemize}
    \item Are the questions for which the models scored poorly just too difficult or unanswerable?
    \item Are there areas in which the performance of humans is very different from the ones of the models?
\end{itemize}
To answer the first question we selected X questions that all \glspl{llm} (model names) from an initial scoring round did not answer correctly.
From those we picked X diverse one using greedy MaxMin sampling on the embeddings on the questions computed using BART (see below).

\paragraph{Study design}
For our initial study we wanted to maximize the response rate given our available resources. For this reason, we did not opt for a highly controlled study setting. That is, while users were prompted to not use external tools other than a calculator and to not consult with other humans, we do not have any way to verify that the participants complied with those rules. Note that users were also allowed to skip questions.

@@

@@

\appendix

\begin{figure}
    \centering
    \includegraphics{figures/flesch_kincaid_reading_ease.pdf}
    \caption{Example question.}
    \label{fig:flesch_kincaid_reading_ease}
\end{figure}

\subsection{Parsing verification}\label{sec:manually-verified-parsing}
For validating the parsing workflow, we randomly sampled four questions per topic and manually verified that the completions of the model were parsed correctly.

\subsection{Model performance}

\begin{figure}
    \centering
    \includegraphics{figures/all_questions_models_completely_correct_radar.pdf}
    \caption{Caption}
    \label{fig:all_questions_models_completely_correct_radar}
    \script{analyze_model_reports.py}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{figures/human_timing.pdf}
    \script{analyze_human_data.py}
    \label{fig:human_timing}
    \caption{}
\end{figure}

Interestingly, we found no significant correlation between the experience of the human scorers and the performance on the questions (Spearman's \(\rho \approx \variable{output/spearman_\experience_score.txt}\), \(p \approx \variable{output/spearman_\experience_score_p.txt}\)).

\begin{figure}
    \centering
    \includegraphics{figures/experience_vs_correctness.pdf}
    \script{analyze_human_data.py}
    \label{fig:experience_vs_correctness}
    \caption{}
\end{figure}

>>>>

\section{Introduction}
\Glspl{llm} are \gls{ml} models trained on massive amounts of text to complete sentences. 
Aggressive scaling of the models has led to a rapid increase in their capabilities,\cite{brown2020language} with the leading models now being able to pass the BAR exam or identify and autonomously buy chemicals when augmented with external tools such as web search and synthesis planners.\cite{openai2024gpt4}
While some see in them \enquote{sparks of \gls{agi}},\cite{bubeck2023sparks} others consider them as \enquote{stochasitc parrots}---i.e., systems that only regurgitate what they have been trained on.\cite{bender2021dangers}
In any case, the promise of so-called foundation models (the leading ones being called frontier models) is that they show the ability solve a wide variety of tasks they have not been explicitly trained on.\cite{bommasani2021opportunities, anderljung2023frontier}
This led to tremendous economic interest and investment in those generative models, with an expected market of more than \$1.3 trillion (almost \$30 billion for drug discovery applications) by 2032.\cite{bloomberg}

Chemists and materials scientists have quickly caught on the mounting attention given to \glspl{llm}, with some voices even suggesting that \enquote{the future of chemistry is language}.\cite{White_2023}
This statement is motivated by a growing number of reports that use \glspl{llm} to properties of molecules or materials,\cite{jablonka202314, jablonka2024leveraging, xie2024fine} to optimize reactions\cite{ramos2023bayesian, kristiadi2024sober} or generate materials,\cite{rubungo2023llm, flam2023language, gruver2024fine} extract information,\cite{Patiny_2023, Dagdelen_2024, Zheng_2024, lála2023paperqa, caufield2023structured} or to even build \enquote{autonomous} systems that can autonomously perform reactions in the physical world based on commands provided in natural language.\cite{bran2023chemcrow, Boiko_2023, darvish2024organa}
But even more than that, since a lot---if not most---of the information about chemistry is currently stored and communicated in text there is a strong reason to believe that there is a lot of untapped potential in \glspl{llm} for chemistry and materials science.
For instance, most insights in chemical research does not directly originate from data stored in databases but rather from the scientists and their ability to interpret data. 
Most of these insights are in form of text in scientific publications. 
Thus, operating on such text might be our best way of unlocking and \emph{learning} from these insights.
This might ultimately lead to copilot systems for chemists that can provide answers to questions or even suggest new experiments based on vastly more information than a human could ever read.
Such a usage mode is especially interesting in the face of recent advanced in autonomous laboratories.\cite{Boiko_2023, bran2023chemcrow, darvish2024organa, granda2018controlling, Angello_2022, coley2019robotic, Burger_2020}
However, the rapid increase in capabilities of chemical \gls{ml} models led (even before the recent interest in \glspl{llm}) to concerns about the potential for dual use of these technologies, e.g., for the design of chemical weapons.\cite{gopal2023releasing, ganguli2022red, Urbina_2022, campbell2023censoring, moulange2023towards}
To some extent, this is not surprising as any technology, that, for instance, can predict the toxicity of molecules (to be used in drug discovery) can also be used to predict the toxicity of molecules (to be used in chemical warfare).

It is important to realize that the user base of such models is not only limited to experts in chemistry and materials science that reflect every output such models produce. 
For example, many students frequently consolidate those tools---perhaps even to prepare chemical experiments.\cite{Intelligent.com_2023}
This also implies to users from the general public, who might consider using \glspl{llm} to answer questions about the safety of chemicals.
Thus, for some users, misleading information---especially about safety-related aspects---might lead to harmful outcomes. 
However, even for experts the chemical understanding and reasoning capabilities are important as they will determine the capabilities and limitations of their models in their work, e.g., in copilot systems for chemists.
Unfortunately, apart from anecdotal reports there is little evidence on how \glspl{llm} perform compared to experts.

Thus, to better understand what \glspl{llm} can do for chemistry and materials science, and where they might be improved with further developments, evaluation frameworks are needed to allow us to systematically measure progress and mitigate potential harms.
For the development of \glspl{llm}, evaluation is currently mostly performed via standardized benchmarks suites such as BigBench\cite{srivastava2022beyond} or the LM Eval Harness.\cite{eval-harness}
The former contains, among 204 tasks (such as linguistic puzzles), only two tasks classified as \enquote{chemistry related} whereas the latter contains no specific chemistry tasks.
Due to the lack of widely excepted standard benchmarks, the developers of chemical language models\cite{jablonka2024leveraging, guo2023large, ahmad2022chemberta2, Cai_2024, frey2023neural} frequently utilize language-interfaced\cite{dinh2022lift} tabular datasets such as the ones reported in MoleculeNet,\cite{wu2018moleculenet} Therapeutic Data Commons\cite{huang2021therapeutics} or MatBench.\cite{dunn2020benchmarking}
In these cases, the models are  evaluated on predicting very specific properties of molecules (e.g., solubility, toxicity, or reactivity) or on predicting the outcome of specific chemical reactions.
This, however, only gives a very limited view of the general chemical capabilities of the models.

While some benchmark based on university entrance exams\cite{Zaki_2024, arora2023llms} or automatic text mining\cite{song2023honeybee, wei2021chemistryqa} have been proposed, none of them have been widely accepted.
This is likely because they cannot automatically be used with black box (or tool-augmented) systems, do not cover a wide range of topics, or are not carefully validated by experts.
On top of that, the existing benchmarks are not designed to be used with models that support special treatment of molecules or equations and do not provide insights in how the models compare relative to experts.

In this work, we report a novel benchmarking framework, which we call chembench, and use it to reveal limitations of current frontier models for the use in the chemical sciences.
Our benchmark consists of \variable{output/total_number_of_questions.txt}\unskip question answer pairs manually (\variable{output/manually_generated.txt}\unskip) or semi-automatically (\variable{output/automatically_generated.txt}\unskip) compiled from diverse sources.
Our corpus large fraction of the topics taught in undergraduate and graduate chemistry curricula and can be used to evaluate any system that can return text (i.e., also an tool-augmented systems).

To contextualize the scores, we also surveyed more than \variable{output/number_experts.txt} experts in chemistry (for a total of number of more than \variable{output/total_hours.txt}\unskip) on a subset of the benchmark corpus to be able to compare the performance of current frontier models with the one of humans.
Our results indicate that current frontier models perform \enquote{superhuman} on some aspects of chemistry but in many cases, included safety-related ones, might be very misleading.
The results indicate the limitations that current models still need to overcome to find use in autonomous systems for chemists and underline the importance of careful evaluation of the capabilities of \glspl{llm} in the chemical sciences.

@@

To compile our benchmark corpus we utilized a broad list of sources (see \Cref{sec:curation}), ranging from university exams to semi-automatically generated questions based on curated subsets of data in chemical databases.
For quality assurance, all questions have been reviewed by at least one scientist in addition to the original curator.

Importantly, our large pool of questions encompasses a wide range of topics.
This can be seen, for example, in \Cref{fig:topic_barplot} in which we compare the number of questions in different subfields of the chemical sciences (see \Cref{sec:meth-topic} for details on how we assigned topics).
The distribution of topics is also evident from \Cref{fig:pca} in which we visualize the questions in a two-dimensional space using a \gls{pca} on the embeddings of the questions.
In this representation, semantically similar questions are close to each other, and we color the points based on a classification into XX topics. 
It is clear that a focus of chembench (by design) lies on safety-related aspects which in \Cref{fig:pca} appear as large distinct clusters.

@@

While many of the existing benchmarks are designed around \gls{mcq}, this does not reflect the reality of chemistry education and research.
For this reason, chembench samples both \gls{mcq} and open-ended questions in a balanced way (\variable{output/mcq_questions.txt} \gls{mcq} questions vs. \variable{output/non_mcq_questions.txt} open-ended questions).
As one might expect, most questions are difficult to read according to the Flesch–Kincaid readability test (\variable{output/flesch_kincaid_reading_ease.txt}).\cite{kincaid1975derivation}

\paragraph{\enquote{Tiny} subset}
It is important to note that for routine evaluations, a smaller subset of the corpus might be more practical.\cite{polo2024tinybenchmarks}
For instance,~\citet{liang2023holistic} report costs of more than 10,000 USD for \gls{api} calls for a single evaluation on the \gls{helm} benchmark on a model. 
To address this, we also provide a subset (\variable{output/num_tiny_questions.txt} questions) of the corpus that was curated to be a diverse and representative subset of the full corpus (see \Cref{sec:subset-selection} for details on the curation process).
We also used this subset to seed the app we used for the human baseline study.

@@

% \subsection{Future work}
% - Advanced prompting techniques such as CoT and test time strategies

% - More advanced systems with chemnistry specific tools

% - More systematic human baseline: Ask people to come in, do not allow to skip questions

@@

@@

\subparagraph{Oxidation states}

\subparagraph{Electron counts}

@@

To maintain the consistency with the training, we employ distinct prompt templates in general tailored for completion models and instruction-tuned models. 
We pose instruction or impose constraints on the models within these templates to receive response in a specific format so that a robust, fair and consistent parsing can be performed as explained in \Cref{sec:parsing}.
Certain models are trained with special annotations and LaTeX syntax for scientific notations, chemical reactions, or symbols embedded within the text. 
For example, all the SMILES representations are encapsulated within \texttt{[START\_SMILES][\text backslash END\_SMILES]} in Galactica\cite{taylor2022galactica}.
Our prompting strategy consistently adheres to  these details in a model-specific manner by post-processing \LaTeX syntax, chemical symbols, chemical equations and physical units (by either adding or removing wrappers).

\paragraph{Models}
\subparagraph{Completion models}
\texttt{Galactica-120b}\cite{taylor2022galactica}

@@

@@

\paragraph{Question selection} \label{sec:subset-selection}
Since we anticipated that we will not be able to collect enough responses for every question to allow for a meaningful statistical analysis, we decided on showing a relevant subset of all questions to the human scorers.
For selecting the subset, we decided on addressing two questions:
\begin{itemize}
    \item Are the questions for which the models scored poorly just too difficult or unanswerable?
    \item Are there areas in which the performance of humans is very different from the ones of the models?
\end{itemize}
To answer the first question we selected X questions that all \glspl{llm} (model names) from an initial scoring round did not answer correctly.
From those we picked X diverse one using greedy MaxMin sampling on the embeddings on the questions computed using BART (see below).

\paragraph{Study design}
For our initial study we wanted to maximize the response rate given our available resources. 
For this reason, we did not opt for a highly controlled study setting. 
hat is, while users were prompted to not use external tools other than a calculator and to not consult with other humans, we do not have any way to verify that the participants complied with those rules. 
Note that users were also allowed to skip questions.

@@

\footnotesize
\insertcredits
\normalsize

@@

\appendix
\input{appendix.tex}
